Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-07-07 manuscript/pnas_style/manuscript.tex
removed supp materials links

% TODO
% - divide through the temp block supp. figure by combinatorics
% - mention PCA

@@

% A long-standing problem in evolutionary biology is to understand the processes
% that shape the genetic composition of populations. In a population without
% migration, the two processes that change allele frequencies are selection,
% which increases beneficial alleles and removes deleterious ones, and genetic
% drift which randomly changes frequencies as some parents contribute more or
% less alleles to the next generation. Previous efforts to disentangle these
% processes have used genomic samples from a single timepoint and models of how
% selection affects neighboring sites (linked selection). Here, we use genomic
% data taken through time to quantify the contributions of selection and drift to
% genome-wide frequency changes. We show selection acts over short timescales in
% three evolve-and-resequence studies and has a sizeable genome-wide impact.

@@

%\section{Introduction}

@@

We first analyzed the dataset of Barghi et al. (2019), an
evolve-and-resequence study with ten replicate populations exposed to a high
temperature lab environment and evolved for 60 generations, and sequenced every
ten generations. Using the seven timepoints and ten replicate populations, we
estimated the genome-wide $6 \times 6$ temporal covariance matrix $\mathbf{Q}$
for each of the ten replicates. Each row of these matrices represent the
temporal covariance $\cov(\Delta_{_{10}} p_s, \Delta_{_{10}} p_t)$, between the
allele frequency change (in ten-generation intervals, denoted $\Delta_{_{10}}
p_t$) of some initial reference generation $s$ (the row of the matrix), and
some later timepoint $t$ (the column of the matrix). We corrected these
matrices for biases created due to sampling noise, and normalized the entries
for heterozygosity (see SI Appendix, sections \ref{supp:ind-depth-var-corr} and
\ref{supp:matrix-correction}). These covariances are expected to be zero when
only drift is acting, as only heritable variation for fitness can create
covariance between allele frequency changes in a closed population
\cite{Buffalo2019-io}.  Averaging across the ten replicate temporal covariances
matrices, we find temporal covariances that are statistically significant (95\%
block bootstraps CIs do not contain zero), consistent with linked selection
perturbing genome-wide allele frequency changes over very short time periods.
The covariances between all adjacent time intervals are positive and then decay
towards zero as we look at more distant time intervals (Figure
\ref{fig:figure-1} A), as expected when directional selection affects linked
variants' frequency trajectories until ultimately linkage disequilibrium and
the associated additive genetic variance for fitness decays (which could occur
as a population reaches a new optimum, and directional selection weakens)
\cite{Buffalo2019-io}. The temporal covariances per replicate are noisier but
this general pattern holds; see SI Appendix, Fig.
\ref{suppfig:barghi-cov-panels}.

Since our covariances are averages over loci, the covariance estimate could be
strongly affected by a few outlier regions. To test whether large outlier
regions drive the genome-wide signal we see in the Barghi et al. (2019) data,
we calculate the covariances in 100kb windows along the genome (we refer to
these as windowed covariances throughout) and take the median windowed
covariance, and trimmed-mean windowed covariance, as a measure of the
genome-wide covariance robust to large-effect loci. These robust estimates (SI
Appendix, Table \ref{supp:table-trimmed-mean} and SI Appendix, Fig.
\ref{suppfig:barghi-trimmed-mean}) confirm the patterns we see using the mean
covariance, establishing that genomic temporal covariances are non-zero
due to the impact of selection acting across many genomic regions.

@@

\begin{align}
  G(t) = \frac{ \sum_{i=0}^{t-1} \sum_{j \ne i}^{t-1} \cov(\Delta p_i, \Delta p_j)}{\var(p_t - p_0)}.
\end{align}
%
This estimates the impact of selection on allele frequency change between
the initial generation $0$ and some later generation $t$, which can be varied
to see how this quantity grows through time. When the sum of the covariances
is positive, this measure can intuitively be understood as a lower bound on
relative fraction of allele frequency change normally thought of as ``drift''
that is actually due to selection. Additionally, $G(t)$ can be understood as a
short-timescale estimate of the reduction in neutral diversity due to linked
selection (or equivalently the reduction in neutral effective population size
needed to account linked selection, see SI Appendix, section
\ref{supp:reduction-approx}). Since the Barghi et al. (2019) experiment is
sequenced every ten generations, the numerator uses the covariances estimated
between ten-generation blocks of allele frequency change; thus the strong,
unobservable, covariances between adjacent generations do not contribute to the
numerator of $G(t)$. Had these covariances been measurable on shorter
timescales, their cumulative effect would likely have been higher yet (see SI
Appendix, sections \ref{supp:barghi-covs} and \ref{supp:tempblocks} for more
details).  Additionally, selection inflates the variance in allele
frequency change per generation; however, this effect cannot be easily
distinguished from drift. For both these reasons, our measure $G(t)$ is quite
conservative (we demonstrate this through simulations in SI Appendix, section
\ref{supp:tempblocks}).  Still, we find a remarkably strong signal. Greater
than $20\%$ of total, genome-wide allele frequency change over 60 generations
is the result of selection (Figure \ref{fig:figure-1} B). This proportion of
variance attributable to selection builds over time in Figure
\ref{fig:figure-1}B as the effects of linked selection are compounded over the
generations unlike genetic drift. Our G(t) starts to plateau to a constant
level as the covariances from earlier generations have decayed and so no longer
contribute as strongly (Figure \ref{fig:figure-1}).

Additionally, we looked for a signal of temporal autocovariance in Bergland et
al. (2014), a study that collected \emph{Drosophila melanogaster} through
Spring-Fall season pairs across three years. If there was a strong  pattern of
genome-wide fluctuating selection, we might expect a pattern of positive
covariances between similar seasonal changes, e.g.  Spring-Fall in two adjacent
years, and negative covariances between dissimilar seasonal changes, e.g.
Spring-Fall and Fall-Spring in two adjacent years.  However, we find no such
signal over years, and in reproducing their original analysis, we find that
their number of statistically significant seasonal SNPs is not enriched
compared to an empirical null distribution created by permuting seasonal
labels; we discuss this in more depth in SI Appendix, section
\ref{supp:bergland-reanalysis}.

@@

A benefit of between-replicate covariances is that unlike temporal covariances,
these can be calculated with only two sequenced timepoints and a replicated
study design. This allowed us to assess the impact of linked selection in
driving convergent patterns of allele frequency change across replicate
populations in two other studies. First, we reanalyzed the selection experiment
of Kelly and Hughes (2019), which evolved three replicate wild populations of
\emph{Drosophila simulans} for 14 generations adapting to a novel laboratory
environment. Since each replicate was exposed to the same selection pressure
and share linkage disequilibria common to the original natural founding
population, we expected each of the three replicate populations to have
positive convergence correlations.  We find all three convergent
correlation coefficients between replicate pairs are significant (Figure
\ref{fig:figure-2}B), and average to 0.36 ($95\%$ CI $[0.31, 0.40]$).
Additionally, we can calculate the proportion of the total variance in
allele frequency change from convergent selection pressure, analogous to
our $G(t)$, where the numerator is the convergent covariance and the
denominator is the total variance (see SI appendix, section
\ref{supp:replicate-g}). We find that 37\% of the total variance is due to
shared allele frequency changes caused by selection (95\% CI [29\%, 41\%];
these are similar to the convergence correlation, since the variance is
relatively constant across the replicates.

@@

The presence of an unselected control line provides an alternative way to
partition the effects of linked selection and genetic drift: we can compare the
total variance in allele frequency change of the control line (which excludes
the effect of artificial selection on allele frequencies) to the total variance
in frequency change of the Longshanks selection lines. This allows us to
estimate the increase in variance in allele frequency change due to selection,
which we can further partition into the effects of selection shared between
selection lines and those unique to a selection line by estimating the shared
effect through the observed covariance between replicates (see Materials and
Methods \ref{sec:mm-partition} and SI Appendix, section
\ref{supp:replicate-g} for more details).  We estimate at least 32\% (95\% CI
$[21\%, 48\%]$) of the variance in allele frequency change is driven by the
effects of selection, of which 14\% (95\% CI $[3\%, 33\%]$) is estimated to be
unique to a selection line, and 17\% (95\% CI $[9\%, 23\%]$) is the effect of
shared selection between the two Longshanks selection lines.

% We can partition the
% allele frequency change between the two timepoints (20 generations apart) for a
% Longshanks line as $\Delta p_{t,\mathrm{LS1}} = \Delta_{_{D}}
% p_{t,\mathrm{LS1}} + \Delta_{_{U}} p_{t,\mathrm{LS1}} + \Delta_{_S}
% p_{t,\mathrm{LS}}$ where these terms are the decomposition in the allele
% frequency change due to drift in Longshanks replicate 1 ($\Delta_{_D}
% p_{t,\mathrm{LS1}}$), selection unique to the LS1 replicate ($\Delta_{_U}
% p_{t,\mathrm{LS1}}$), and selection response shared between the two Longshanks
% replicates ($\Delta_{_S} p_{t,\mathrm{LS}}$) respectively (and similarly for
% the Longshanks two line, LS2). By construction we will assume that each of
% these terms are uncorrelated within replicates, and that only the shared term
% covaries between the replicates.  Assuming that we can approximate the
% contribution of genetic drift in the LS lines as the variance in allele
% frequency change observed in the control, i.e.  $\var(\Delta
% p_{t,\mathrm{Ctrl}}) = \var(\Delta_{_{D}}
% p_{t,\mathrm{LS2}})=\var(\Delta_{_{D}} p_{t,\mathrm{LS2}})$, then we can
% estimate the increase in variance in allele frequency change due to selection
% as $(\var(\Delta p_{t,\mathrm{LS1}}) + \var(\Delta p_{t,\mathrm{LS2}}))/2 -
% \var(\Delta p_{t,\mathrm{Ctrl}})$ and the shared effect of selection across
% selected lines as $\cov(\Delta p_{t,\mathrm{LS1}}, \Delta p_{t,\mathrm{LS2}})$

@@

  \caption{\footnotesize {\bf A}, {\bf B}: The distribution of temporal
    covariances calculated in 100kb genomic windows from the
    Barghi et al. (2019) study, plotted alongside an empirical neutral null
    distribution created by recalculating the windowed covariances on 1,000
    sign permutations of allele frequency changes within tiles. The number of histogram
    bins is 88, chosen by cross validation (SI Appendix, section
    \ref{suppfig:barghi-cross-validation-binsize}). In subfigure {\bf A},
    windowed covariances $\cov(\Delta p_t, \Delta p_{t+k})$ are separated by
    $k=2 \times 10$ generations and in subfigure {\bf A} the covariances are
    separated by $k=4 \times 10$ generations; each $k$ is an off-diagonal from
    the variance diagonal of the temporal covariance matrix (see cartoon of
    upper-triangle of covariance matrix in subfigures {\bf A} and {\bf B},
    where the first diagonal is the variance, and the dark gray indicates which
    off-diagonal of the covariance matrix is plotted in the histograms). {\bf
    C}: The lower and upper tail probabilities of the observed windowed
    covariances, at 20\% and 80\% quintiles of the empirical neutral null
    distribution, for varying time between allele frequency changes (i.e. which
    off-diagonal $k$). The confidence intervals are  $95\%$ block-bootstrap
  confidence intervals, and the light gray dashed line indicates the 20\% tail
probability expected under the neutral null. Similar figures for different
values of $k$ are in SI Appendix, section
\ref{suppfig:barghi-empnull-tilecovs}. }

@@

We observed that in the longest study  we analyzed \cite{Barghi2019-qy},
some genome-wide temporal covariances become negative at future timepoints (see
the first two rows in Figure \ref{fig:figure-1}A). This shows that alleles that
were on average going up initially are later going down in frequency, i.e.
that the average direction of selection experienced by alleles has flipped.
This might reflect either a change in the environment or the genetic
background, due to epistatic relationships among alleles altered by frequency
changes (which can occur during an optima shift; \cite{Hayward2019-kt}) or
recombination breaking up selective alleles. Such reversals in selection
dynamics could be occurring at other timepoints but the signal of a change in
the direction of selection at particular loci may be washed out when we
calculate our genome-wide average temporal covariances.  To address this
limitation, we calculated the distribution of the temporal covariances over
100kb windowed covariances (Figure \ref{fig:figure-3} shows these distributions
pooling across all replicates; see SI Appendix, Fig.
\ref{suppfig:barghi-offset-replicate-panels} for individuals replicates). The
covariance estimate of each genomic window will be noisy, due to sampling and
genetic drift, and the neutral distribution of the covariance is complicated
due to linkage disequilibria, which can occur over long physical distances in
E\&R and selection studies \cite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}.  To
address this, we have developed a permutation-based procedure that constructs
an empirical neutral null distribution by randomly flipping the sign of
the allele frequency changes in each genomic window (i.e. a single random sign
flip is applied to all loci in a window). This destroys the systematic
covariances created by linked selection and creates a sampling distribution of
the covariances spuriously created by neutral genetic drift while preserving
the complex dependencies between adjacent loci created by linkage
disequilibrium.  This empirical neutral null distribution is conservative in
the sense that the variances of the covariances are wider than expected under
drift alone, as selection not only creates covariance between time
intervals, but also inflates the magnitude of allele frequency change within a
time-interval.  We see (Figure \ref{fig:figure-3} A and B) that there are an
empirical excess of windows with positive covariances between close timepoints
compared to the null distribution (a heavier right tail), and that this then
shifts to an excess of windows with negative covariances between more distant
timepoints (a heavier left tail).

We quantified the degree to which the left and
right tails are inflated compared to the null distribution as a function of
time, and see excesses in both tails in Figure \ref{fig:figure-3}C. This
finding is also robust to sign-permuting allele frequency changes on a
chromosome-level, the longest extent that gametic linkage disequilibria can
extend (SI Appendix, Fig. \ref{suppfig:barghi-tailprobs-seqid}). We see a
striking pattern that the windowed covariances not only decay towards zero, but
in fact become negative through time, consistent with many regions in the
genome having had a reversed fitness effect at later timepoints.

@@

  \caption{Forward-in-time simulations demonstrate how temporal covariance,
    $G(t)$ trajectories, and convergence correlations arise during optima
    shifts of two different magnitudes, under Gaussian stabilizing selection.
    ({\bf A}) Trait means across 30 replicate before and after optima shifts
    (solid lines), for two different magnitudes (indicated by color). The
    optimal trait values are indicated by the purple and yellow dashed lines.
    ({\bf B}) Mean temporal covariance $\cov(\Delta p_5, \Delta p_t)$ across 30
    simulation replicates, where $t$ varies along the x-axis (points), with a
    loess-smoothed average (solid line). ({\bf C}) $G(t)$ trajectories through
    time, for 30 replicate simulations across two optima shifts. The solid line
    is a loess-smoothed average. ({\bf D}) The convergence correlations between
    two populations (each $1000$ diploids) split from a common population, that
    underwent either an optima shift in the same direction (converge) and
    opposite directions (diverge) at generation five. In subfigures ({\bf B}),
    ({\bf C}), and ({\bf D}), directional selection begins at generation five,
    when the optima shifts; this is indicated by the vertical dashed red line
    (see SI Appendix Section \ref{supp:tempcov-gss} for details on these
    simulations).}

@@

Finally we used forward-in-time simulations to explore the conditions under
which temporal and convergent correlations arise. We show a subset of our
results for a model of stabilizing selection on a phenotype where directional
selection is induced by a sudden shift in the optimum phenotype of varying
magnitudes (Figure \ref{fig:figure-4}A). We find that positive temporal
covariances are produced by such selection (Figure \ref{fig:figure-4}B), and
that these positive temporal covariances can compound together to generate a
large proportion of allele frequency change being due to selection (i.e. large
$G(t)$) over the relatively short time periods similar to our analyzed
selection datasets span (Figure \ref{fig:figure-4}C).  The magnitude of
$G(t)$ increases with the strength of selection, i.e. the variance in
fitness, such that stronger selection generates larger proportions of allele
frequency change.  We find a similar picture of stronger convergent selection
pressures generating larger convergence correlations (Figure
\ref{fig:figure-4}D; see also SI Appendix, Fig.
\ref{suppfig:convergence-corrs} for how other factors impact convergence
correlations).

Averaging across replicates, these simulation results show $G(t)$ is relatively
insensitive to the number of loci underlying the trait. However, if only a
small number of loci influence the trait, the $G(t)$ trajectories are typically
much more stochastic \emph{across} replicates. This reaffirms that the
genome-wide linked selection response we see in the Barghi et al.  (2019) data
is highly polygenic (compare Figure \ref{fig:figure-1}B to SI Appendix, Fig.
\ref{suppfig:sim-expfit-Gs}). Furthermore, using our simulations we find
that sampling only every 10 generations does indeed mean that our estimates of
$G(t)$ are an underestimate of the proportional effect of linked selection
as they cannot include the covariance between closely spaced generations (see
SI Appendix, Fig.  \ref{suppfig:supp-blocks}).

Additionally, we explored other modes of selection with simulations.
We find that the long term dynamics of the covariances under directional
truncation selection, which generates substantial epistasis, are richer than we
see under Gaussian Stabilizing Selection (GSS) and multiplicative selection (SI
Appendix, Fig.  \ref{suppfig:supp-trunc}). We also conducted simulations of
purifying selection alone (i.e. background selection) and find that this can
also generate positive temporal covariances (SI Appendix, Fig.
\ref{suppfig:supp-bgs-covs}) and under some circumstances, can even generate
convergence correlations (SI Appendix, Fig.  \ref{suppfig:supp-bgs-converg}).
Thus it is unlikely that the signatures of linked selection we see are entirely
the result of the novel selection pressure the populations are exposed to, and
some of this signature may be ongoing purifying selection. Only in the
case of the Longshanks experiment, does the control line allow us to conclude
that selection that is almost entirely due to the novel selection
pressure.

@@

To remove systematic covariances in allele frequency change caused by tracking
the reference or minor allele, we randomly choose an allele to track
frequency for each locus. Then, we calculate the variance-covariance
matrix of allele frequency changes using a Python software package we have
written, available at \url{http://github.com/vsbuffalo/cvtk}. This
simultaneously calculates temporal variances and covariances, and replicate
covariances and uses the sampling depth and number of diploid individuals to
correct for bias in the variance estimates and a bias that occurs in covariance
estimates between adjacent timepoints due to shared sampling noise (see
SI Appendix, sections \ref{supp:ind-depth-var-corr},
\ref{supp:cov-corr}, and \ref{supp:matrix-correction} for mathematical details
of these estimators). We assess that our bias correction procedure is working
adequately through a series of diagnostic plots that ensure that the procedure
removes the relationship between sampling depth and uncorrected variance and
covariances (SI Appendix, Fig. \ref{suppfig:bergland-correction}).  Through
our simulations we find that our estimates can differ based on how fixations
and losses are handled in long time-series (SI Appendix, section 
\ref{supp:fixation}) but none of our findings in the main text are
qualitatively altered by this decision (SI Appendix, Figs.
\ref{suppfig:supp-fig-1-nofix} and \ref{suppfig:supp-fig-3-nofix}).

@@

To infer the uncertainty of covariance, convergence correlation, and $G(t)$
estimates, we used a block bootstrap procedure. This bootstrap procedure
resamples blocks of loci, rather than individual loci, to infer the
uncertainty of a statistic in the presence of unknown correlation between loci.
As most estimators in this paper are ratios (e.g. covariance standardize by
sample heterozygosity, $G(t)$, and the convergence correlation), which we
estimate with a ratio of averages, we exploit the linearity of expectation for
efficient computation of bootstrap samples (see SI Appendix, Fig.
\ref{supp:block-bootstrap} for details).

@@

To explore how aspects of genetic architecture, models of selection, and
experimental design impact temporal covariance, the $G(t)$ trajectories, and
convergence correlations, we ran extensive forward-in-time simulations
using SLiM \cite{Haller2019-vu}; here we discuss the Gaussian Stabilizing
Selection simulations in Figured \ref{fig:figure-4}, but SI Appendix, section
\ref{supp:forward} describes these simulation routines and others in detail.

>>>>

@@

@@

@@

We first analyzed the dataset of Barghi et al. (2019), an evolve-and-resequence
study with ten replicate populations exposed to a high temperature lab
environment and evolved for 60 generations, and sequenced every ten
generations. Using the seven timepoints and ten replicate populations, we
estimated the genome-wide $6 \times 6$ temporal covariance matrix $\mathbf{Q}$
for each of the ten replicates. Each row of these matrices represent the
temporal covariance $\cov(\Delta_{_{10}} p_s, \Delta_{_{10}} p_t)$, between the
allele frequency change (in ten-generation intervals, denoted $\Delta_{_{10}}
p_t$) of some initial reference generation $s$ (the row of the matrix), and
some later timepoint $t$ (the column of the matrix). We corrected these
matrices for biases created due to sampling noise, and normalized the entries
for heterozygosity (see SI Appendix, Sections S1.2 and S1.4). These covariances
are expected to be zero when only drift is acting, as only heritable variation
for fitness can create covariance between allele frequency changes in a closed
population \cite{Buffalo2019-io}.  Averaging across the ten replicate temporal
covariances matrices, we find temporal covariances that are statistically
significant (95\% block bootstraps CIs do not contain zero), consistent with
linked selection perturbing genome-wide allele frequency changes over very
short time periods.  The covariances between all adjacent time intervals are
positive and then decay towards zero as we look at more distant time intervals
(Figure \ref{fig:figure-1} A), as expected when directional selection affects
linked variants' frequency trajectories until ultimately linkage disequilibrium
and the associated additive genetic variance for fitness decays (which could
occur as a population reaches a new optimum, and directional selection weakens)
\cite{Buffalo2019-io}. The temporal covariances per replicate are noisier but
this general pattern holds; see SI Appendix, Fig. S23.

Since our covariances are averages over loci, the covariance estimate could be
strongly affected by a few outlier regions. To test whether large outlier
regions drive the genome-wide signal we see in the Barghi et al. (2019) data,
we calculate the covariances in 100kb windows along the genome (we refer to
these as windowed covariances throughout) and take the median windowed
covariance, and trimmed-mean windowed covariance, as a measure of the
genome-wide covariance robust to large-effect loci. These robust estimates (SI
Appendix, Table S1 and Fig. S24) confirm the patterns we see using the mean
covariance, establishing that genomic temporal covariances are non-zero due to
the impact of selection acting across many genomic regions.

@@

\begin{align}
  G(t) = \frac{ \sum_{i=0}^{t-1} \sum_{j \ne i}^{t-1} \cov(\Delta p_i, \Delta p_j)}{\var(p_t - p_0)}.
\end{align}
%
This estimates the impact of selection on allele frequency change between the
initial generation $0$ and some later generation $t$, which can be varied to
see how this quantity grows through time. When the sum of the covariances is
positive, this measure can intuitively be understood as a lower bound on
relative fraction of allele frequency change normally thought of as ``drift''
that is actually due to selection. Additionally, $G(t)$ can be understood as a
short-timescale estimate of the reduction in neutral diversity due to linked
selection (or equivalently the reduction in neutral effective population size
needed to account linked selection, see SI Appendix, Section S7). Since the
Barghi et al. (2019) experiment is sequenced every ten generations, the
numerator uses the covariances estimated between ten-generation blocks of
allele frequency change; thus the strong, unobservable, covariances between
adjacent generations do not contribute to the numerator of $G(t)$. Had these
covariances been measurable on shorter timescales, their cumulative effect
would likely have been higher yet (see SI Appendix, Sections S2 and
S8.4 for more details).  Additionally, selection inflates the variance in
allele frequency change per generation; however, this effect cannot be easily
distinguished from drift. For both these reasons, our measure $G(t)$ is quite
conservative (we demonstrate this through simulations in SI Appendix, Section
S8.4).  Still, we find a remarkably strong signal. Greater than $20\%$ of
total, genome-wide allele frequency change over 60 generations is the result of
selection (Figure \ref{fig:figure-1} B). This proportion of variance
attributable to selection builds over time in Figure \ref{fig:figure-1}B as the
effects of linked selection are compounded over the generations unlike genetic
drift. Our G(t) starts to plateau to a constant level as the covariances from
earlier generations have decayed and so no longer contribute as strongly
(Figure \ref{fig:figure-1}).

Additionally, we looked for a signal of temporal autocovariance in Bergland et
al. (2014), a study that collected \emph{Drosophila melanogaster} through
Spring-Fall season pairs across three years. If there was a strong  pattern of
genome-wide fluctuating selection, we might expect a pattern of positive
covariances between similar seasonal changes, e.g.  Spring-Fall in two adjacent
years, and negative covariances between dissimilar seasonal changes, e.g.
Spring-Fall and Fall-Spring in two adjacent years.  However, we find no such
signal over years, and in reproducing their original analysis, we find that
their number of statistically significant seasonal SNPs is not enriched
compared to an empirical null distribution created by permuting seasonal
labels; we discuss this in more depth in SI Appendix, Section S6).

@@

A benefit of between-replicate covariances is that unlike temporal covariances,
these can be calculated with only two sequenced timepoints and a replicated
study design. This allowed us to assess the impact of linked selection in
driving convergent patterns of allele frequency change across replicate
populations in two other studies. First, we reanalyzed the selection experiment
of Kelly and Hughes (2019), which evolved three replicate wild populations of
\emph{Drosophila simulans} for 14 generations adapting to a novel laboratory
environment. Since each replicate was exposed to the same selection pressure
and share linkage disequilibria common to the original natural founding
population, we expected each of the three replicate populations to have
positive convergence correlations.  We find all three convergent correlation
coefficients between replicate pairs are significant (Figure
\ref{fig:figure-2}B), and average to 0.36 ($95\%$ CI $[0.31, 0.40]$).
Additionally, we can calculate the proportion of the total variance in allele
frequency change from convergent selection pressure, analogous to our $G(t)$,
where the numerator is the convergent covariance and the denominator is the
total variance (see SI appendix, Section S4). We find that 37\% of the total
variance is due to shared allele frequency changes caused by selection (95\% CI
[29\%, 41\%]; these are similar to the convergence correlation, since the
variance is relatively constant across the replicates.

@@

The presence of an unselected control line provides an alternative way to
partition the effects of linked selection and genetic drift: we can compare the
total variance in allele frequency change of the control line (which excludes
the effect of artificial selection on allele frequencies) to the total variance
in frequency change of the Longshanks selection lines. This allows us to
estimate the increase in variance in allele frequency change due to selection,
which we can further partition into the effects of selection shared between
selection lines and those unique to a selection line by estimating the shared
effect through the observed covariance between replicates (see Materials and
Methods \ref{sec:mm-partition} and SI Appendix, Section S4 for more details).
We estimate at least 32\% (95\% CI $[21\%, 48\%]$) of the variance in allele
frequency change is driven by the effects of selection, of which 14\% (95\% CI
$[3\%, 33\%]$) is estimated to be unique to a selection line, and 17\% (95\% CI
$[9\%, 23\%]$) is the effect of shared selection between the two Longshanks
selection lines.

@@

  \caption{\footnotesize {\bf A}, {\bf B}: The distribution of temporal
    covariances calculated in 100kb genomic windows from the
    Barghi et al. (2019) study, plotted alongside an empirical neutral null
    distribution created by recalculating the windowed covariances on 1,000
    sign permutations of allele frequency changes within tiles. The number of histogram
    bins is 88, chosen by cross validation (SI Appendix, Section
    S25). In subfigure {\bf A},
    windowed covariances $\cov(\Delta p_t, \Delta p_{t+k})$ are separated by
    $k=2 \times 10$ generations and in subfigure {\bf A} the covariances are
    separated by $k=4 \times 10$ generations; each $k$ is an off-diagonal from
    the variance diagonal of the temporal covariance matrix (see cartoon of
    upper-triangle of covariance matrix in subfigures {\bf A} and {\bf B},
    where the first diagonal is the variance, and the dark gray indicates which
    off-diagonal of the covariance matrix is plotted in the histograms). {\bf
    C}: The lower and upper tail probabilities of the observed windowed
    covariances, at 20\% and 80\% quintiles of the empirical neutral null
    distribution, for varying time between allele frequency changes (i.e. which
    off-diagonal $k$). The confidence intervals are  $95\%$ block-bootstrap
  confidence intervals, and the light gray dashed line indicates the 20\% tail
probability expected under the neutral null. Similar figures for different
values of $k$ are in SI Appendix, Section S27.}

@@

We observed that in the longest study  we analyzed \cite{Barghi2019-qy}, some
genome-wide temporal covariances become negative at future timepoints (see the
first two rows in Figure \ref{fig:figure-1}A). This shows that alleles that
were on average going up initially are later going down in frequency, i.e.
that the average direction of selection experienced by alleles has flipped.
This might reflect either a change in the environment or the genetic
background, due to epistatic relationships among alleles altered by frequency
changes (which can occur during an optima shift; \cite{Hayward2019-kt}) or
recombination breaking up selective alleles. Such reversals in selection
dynamics could be occurring at other timepoints but the signal of a change in
the direction of selection at particular loci may be washed out when we
calculate our genome-wide average temporal covariances.  To address this
limitation, we calculated the distribution of the temporal covariances over
100kb windowed covariances (Figure \ref{fig:figure-3} shows these distributions
pooling across all replicates; see SI Appendix, Fig. S26 for individuals
replicates). The covariance estimate of each genomic window will be noisy, due
to sampling and genetic drift, and the neutral distribution of the covariance
is complicated due to linkage disequilibria, which can occur over long physical
distances in E\&R and selection studies
\cite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}.  To address this, we have developed
a permutation-based procedure that constructs an empirical neutral null
distribution by randomly flipping the sign of the allele frequency changes in
each genomic window (i.e. a single random sign flip is applied to all loci in a
window). This destroys the systematic covariances created by linked selection
and creates a sampling distribution of the covariances spuriously created by
neutral genetic drift while preserving the complex dependencies between
adjacent loci created by linkage disequilibrium.  This empirical neutral null
distribution is conservative in the sense that the variances of the covariances
are wider than expected under drift alone, as selection not only creates
covariance between time intervals, but also inflates the magnitude of allele
frequency change within a time-interval.  We see (Figure \ref{fig:figure-3} A
and B) that there are an empirical excess of windows with positive covariances
between close timepoints compared to the null distribution (a heavier right
tail), and that this then shifts to an excess of windows with negative
covariances between more distant timepoints (a heavier left tail).

We quantified the degree to which the left and right tails are inflated
compared to the null distribution as a function of time, and see excesses in
both tails in Figure \ref{fig:figure-3}C. This finding is also robust to
sign-permuting allele frequency changes on a chromosome-level, the longest
extent that gametic linkage disequilibria can extend (SI Appendix, Fig. S29).
We see a striking pattern that the windowed covariances not only decay towards
zero, but in fact become negative through time, consistent with many regions in
the genome having had a reversed fitness effect at later timepoints.

@@

  \caption{Forward-in-time simulations demonstrate how temporal covariance,
    $G(t)$ trajectories, and convergence correlations arise during optima
    shifts of two different magnitudes, under Gaussian stabilizing selection.
    ({\bf A}) Trait means across 30 replicate before and after optima shifts
    (solid lines), for two different magnitudes (indicated by color). The
    optimal trait values are indicated by the purple and yellow dashed lines.
    ({\bf B}) Mean temporal covariance $\cov(\Delta p_5, \Delta p_t)$ across 30
    simulation replicates, where $t$ varies along the x-axis (points), with a
    loess-smoothed average (solid line). ({\bf C}) $G(t)$ trajectories through
    time, for 30 replicate simulations across two optima shifts. The solid line
    is a loess-smoothed average. ({\bf D}) The convergence correlations between
    two populations (each $1000$ diploids) split from a common population, that
    underwent either an optima shift in the same direction (converge) and
    opposite directions (diverge) at generation five. In subfigures ({\bf B}),
    ({\bf C}), and ({\bf D}), directional selection begins at generation five,
    when the optima shifts; this is indicated by the vertical dashed red line
    (see SI Appendix Section S8.2 for details on these
    simulations).}

@@

Finally we used forward-in-time simulations to explore the conditions under
which temporal and convergent correlations arise. We show a subset of our
results for a model of stabilizing selection on a phenotype where directional
selection is induced by a sudden shift in the optimum phenotype of varying
magnitudes (Figure \ref{fig:figure-4}A). We find that positive temporal
covariances are produced by such selection (Figure \ref{fig:figure-4}B), and
that these positive temporal covariances can compound together to generate a
large proportion of allele frequency change being due to selection (i.e. large
$G(t)$) over the relatively short time periods similar to our analyzed
selection datasets span (Figure \ref{fig:figure-4}C).  The magnitude of $G(t)$
increases with the strength of selection, i.e. the variance in fitness, such
that stronger selection generates larger proportions of allele frequency
change.  We find a similar picture of stronger convergent selection pressures
generating larger convergence correlations (Figure \ref{fig:figure-4}D; see
also SI Appendix, Fig. S12 for how other factors impact convergence
correlations).

Averaging across replicates, these simulation results show $G(t)$ is relatively
insensitive to the number of loci underlying the trait. However, if only a
small number of loci influence the trait, the $G(t)$ trajectories are typically
much more stochastic \emph{across} replicates. This reaffirms that the
genome-wide linked selection response we see in the Barghi et al.  (2019) data
is highly polygenic (compare Figure \ref{fig:figure-1}B to SI Appendix, Fig.
S6). Furthermore, using our simulations we find
that sampling only every 10 generations does indeed mean that our estimates of
$G(t)$ are an underestimate of the proportional effect of linked selection
as they cannot include the covariance between closely spaced generations (see
SI Appendix, Fig. S14).

Additionally, we explored other modes of selection with simulations.  We find
that the long term dynamics of the covariances under directional truncation
selection, which generates substantial epistasis, are richer than we see under
Gaussian Stabilizing Selection (GSS) and multiplicative selection (SI Appendix,
Fig. S18). We also conducted simulations of purifying selection alone (i.e.
background selection) and find that this can also generate positive temporal
covariances (SI Appendix, Fig. S16) and under some circumstances, can even
generate convergence correlations (SI Appendix, Fig. S17).  Thus it is unlikely
that the signatures of linked selection we see are entirely the result of the
novel selection pressure the populations are exposed to, and some of this
signature may be ongoing purifying selection. Only in the case of the
Longshanks experiment, does the control line allow us to conclude that
selection that is almost entirely due to the novel selection pressure.

@@

To remove systematic covariances in allele frequency change caused by tracking
the reference or minor allele, we randomly choose an allele to track frequency
for each locus. Then, we calculate the variance-covariance matrix of allele
frequency changes using a Python software package we have written, available at
\url{http://github.com/vsbuffalo/cvtk}. This simultaneously calculates temporal
variances and covariances, and replicate covariances and uses the sampling
depth and number of diploid individuals to correct for bias in the variance
estimates and a bias that occurs in covariance estimates between adjacent
timepoints due to shared sampling noise (see SI Appendix, Sections S1.2, S1.3,
and S1.4 for mathematical details of these estimators). We assess that our bias
correction procedure is working adequately through a series of diagnostic plots
that ensure that the procedure removes the relationship between sampling depth
and uncorrected variance and covariances (SI Appendix, Fig. S4).  Through our
simulations we find that our estimates can differ based on how fixations and
losses are handled in long time-series (SI Appendix, Section S8.7) but none of
our findings in the main text are qualitatively altered by this decision (SI
Appendix, Figs. S19 and S20).

@@

To infer the uncertainty of covariance, convergence correlation, and $G(t)$
estimates, we used a block bootstrap procedure. This bootstrap procedure
resamples blocks of loci, rather than individual loci, to infer the uncertainty
of a statistic in the presence of unknown correlation between loci.  As most
estimators in this paper are ratios (e.g. covariance standardize by sample
heterozygosity, $G(t)$, and the convergence correlation), which we estimate
with a ratio of averages, we exploit the linearity of expectation for efficient
computation of bootstrap samples (see SI Appendix, Fig. S3 for details).

@@

To explore how aspects of genetic architecture, models of selection, and
experimental design impact temporal covariance, the $G(t)$ trajectories, and
convergence correlations, we ran extensive forward-in-time simulations using
SLiM \cite{Haller2019-vu}; here we discuss the Gaussian Stabilizing Selection
simulations in Figured \ref{fig:figure-4}, but SI Appendix, Section S8
describes these simulation routines and others in detail.
